A Fast and Tight Heuristic for A* in Road Networks

### 2021-01-16 paper/ch_potentials.tex
intro changes

\section{Introduction}
\label{sec:intro}
The past decade has seen a plethora of research on route planning in large street networks~\cite{bdgmpsww-rptn-16}.
Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
Road intersections become nodes.
Road segments become edges.
The weight of an edge is the time a car needs to traverse the road segment.
The problem can be solved with Dijkstra's algorithm~\cite{d-ntpcg-59}.
However, on continental sized networks, it is too slow for many applications.
Thus, speed-up techniques have been developed.
A popular example for such a technique, which has been successfully used in many real world applications, is Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
CH exploit the inherent hierarchy of road networks.
In a preprocessing step, additional shortcut edges are inserted, which allow skipping over unimportant parts of the network during query time.
Another technique, also successfully employed in practice, is Multi-Level-Dijkstra~(MLD)~\cite{dgpw-crprn-13} better known as CRP.
It also exploits the hierarchy of road networks using shortcut edges.
Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.
Their implementation complexity is also quite manageable.
This makes them very attractive in practice.

@@

However, for many real world routing applications, this basic model turns out to be too simple.
For realistic routing, many additional features need to be considered.
This includes but is not limited to turn costs and restrictions, live traffic, user preferences, traffic predictions, temporary driving bans for certain types of vehicles and many more.
Hierarchical speed-up techniques can be extended to support these features.
However, each feature requires significant effort.
Many combinations of specific techniques with certain extensions have their own dedicated research paper(s).
Adding these features introduces significant additional implementation complexity.
Supporting combinations of extensions is at least non-trivial, oftentimes downright impossible.

In this work, we propose a novel building block for routing algorithms which drastically improves on this extensibility problem with very little implementation complexity.
For this, we revisit goal-directed search (A*).
Goal directed search is a simple extension of Dijkstra's algorithm.
An estimate of the remaining distance to the target is used, to change the order in which nodes are traversed during the search.
This guides the search towards the target.
See Figure~\ref{img:search-space} for an example of nodes traversed during a goal-directed search.
This approach is far easier to extend.
Many extensions can be supported without modifying the distance estimation at all.
The downside is that the effectiveness of this approach highly depends on the tightness of the estimate.
There has been extensive research on goal-directed search in the past~\cite{}.
However, on road networks, previous approaches could not deliver performance on par with hierarchical techniques.
We introduce CH-Potentials, a new distance estimation function which can efficiently compute tight distance estimates.
\todo{tight}
CH-Potentials are a simple combination of existing approaches.
CH-Potentials use hierarchical speed-up techniques but only indirectly through goal directed search.
This new building block allows handling extensions orthogonal to the speed-up technique.
\todo{contribution - no better heuristic possible}
% By comparing against Oracle-A*, we show that no other exact two-phase A*-based approach can be significantly faster.
In addition, we propose optimizations for Dijkstra's/A*'s handling of low-degree nodes, common in road networks.

>>>>

\section{Introduction}
\label{sec:intro}
The past decade has seen a plethora of research on route planning in large street networks~\cite{bdgmpsww-rptn-16}.
Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
Nodes represent intersections.
Roads are modeled using edges.
Edges are weighted by their traversal times.
The problem can be solved with Dijkstra's algorithm~\cite{d-ntpcg-59}.
Unfortunately, on continental sized networks, it is too slow for many applications.
Thus, speed-up techniques have been developed.
One popular example are Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
They have been used successfully in many real world applications\todo{citation?}.
A CH exploits the inherent hierarchy of road networks.
In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant roads during query time.
A further popular example is Multi-Level-Dijkstra~(MLD)~\todo{pre CRP zitat?} also known as CRP~\cite{dgpw-crprn-13}.
It is also used in practice~\todo{citation, bing blog?}.
MLD also also uses shortcut edges.
Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.
% --- Unser Argument ist, dass die zu komplex zum erweitern sind... da passen diese 
%     Sätze nicht in die Story
% Their implementation complexity is also quite manageable.    
% This makes them very attractive in practice.

@@

Unfortunately, for many real world applications, this basic graph model is too simplistic.
Many additional features need to be considered.
% --- ich habe turn costs weg gelassen, weil wir die nicht in den experimenten haben, ich will keiner Reviewer auf blöde Ideeen bringen
% --- ich habe temporary driving bans weg gelassen, weil wir das nicht in den Experimenten haben und ich "we can do all of the above features" claimen möchte
This includes turn restrictions, live traffic, user preferences, traffic predictions.%, temporary driving bans for certain types of vehicles.
Specific applications often have additional application-specific requirements.
Extending Dijkstra's algorithm to support these features is usually easy.
On the other hand, while extending hierarchical speed-up techniques is possible, the algorithm development is vastly more complex.
For every feature, dedicated research paper(s) exist that extend CH.
Supporting the combination of several features is even harder. 
For example, we are not aware of any work combining all features mentioned above.
In this paper, we describe an algorithm that handles the combination of all mentioned features.

% --- mit "novel" und "propose" muss man vorsichtig sein, wegen SEA paper von letztem Jahr
We describe an algorithmic building block that mostly decouples feature development from the hierarchical speed-up algorithm.
Our goal is to simplify incorporating new features.
We build upon the A* algorithm \cite{A* literature}.
A* is a goal-directed variant of Dijkstra's algorithm.
It uses a heuristic\todo{heuristic vs potential} to guide the search towards the goal.
A heuristic is function that maps a node $v$ onto a lower bound of the distance from $v$ to the goal.
A*'s running time crucially depends on how tight this lower bound is.
Further, evaluating the heuristic must be fast.
The free flow travel time to the goal is a valid heuristic for most features.
In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
In this paper, we describe CH-Potentials, a technique that achieves both goals.
Internally, the heuristic uses a CH.
Fortunately, this is an implementation detail from the perspective of the A*.
To support a new feature, we only need to modify the A* algorithm.
The heuristic core containing the CH remains untouched.
Extending A* is vastly easier than extending a CH.
This enables us to design an algorithms that supports a multitude of features.
In addition, we describe optimizations for handling of low-degree nodes, common in road networks.
These low degree optimizations are applicable to Dijkstra's algorithm and A*.
